Appendix 4: Results

### 2016-02-12 Harris_2015/manuscript-materials/draft.Rmd
clarify optimization

\noindent where the second sum is over all $\frac{1}{2}n(n-1)$ pairs of $n$
species. In this model, $\alpha_{i}$ is an intercept term determining the amount
that the presence of species $i$ contributes to the log-probability of
$\vec{y}$; it directly controls the prevalence of species $i$. Similarly,
$\beta_{ij}$ is the amount that the co-occurrence of species $i$ and species $j$
contributes to the log-probability; it controls the conditional relationship
between two species, i.e. the probability that they will be found together,
after controlling for the other species in the network (Figure 2A, Figure 2B).
For example, if $\beta_{ij} = +2$, then each species' odds of occurrence would be
$e^2$ times higher when the other one is present (as compared with otherwise
equivalent sites). The relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur. As a result, the model tends---all else
equal---to produce assemblages where many positively-associated species pairs
co-occur and few negatively-associated pairs do (just as an ecologist might
expect). When all else is *not* equal (e.g. Figure 1, where the presence of one
competitor is associated with release from another competitor), then predicting
species' overall co-occurrence rates would require normalizing the relative 
probabilities of possible $\vec{y}$ vectors so that they sum to one (Figure 2B).

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a set of independent $\vec{y}$ vectors
indicating which species are present at each site on the landscape). When
Equation 1 can be normalized, one can find exact maximum likelihood estimates for
$\alpha$ and $\beta$ using numerical optimization. Fully-observed networks like the 
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012], so 
these estimates will represent the global optimum. I used the rosalia package 
[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to 
calculate $p(\vec{y}; \alpha, \beta)$ and its gradients [@murphy_machine_2012]; 
the package passes these functions to the "BFGS" method in R's general-purpose 
optimizer, which then estimates the Markov network parameters.

\noindent \textbf{Simulated landscapes.}
I simulated several sets of landscapes using known parameters to evaluate
different statistical methods' performance. The first set of landscapes included
the three competing species shown in Figure 1. For each of 1000 replicates, I
generated a landscape with 100 sites by sampling from a probability distribution
defined by the figure's interaction coefficients (Appendix 1). Each of the
methods described below was then evaluated on its ability to correctly infer
that the two shrub species competed with one another, despite their frequent
co-occurrence.

>>>>

\noindent where the second sum is over all $\frac{1}{2}n(n-1)$ pairs of $n$
species. In this model, $\alpha_{i}$ is an intercept term determining the amount
that the presence of species $i$ contributes to the log-probability of
$\vec{y}$; it directly controls the prevalence of species $i$. Similarly,
$\beta_{ij}$ is the amount that the co-occurrence of species $i$ and species $j$
contributes to the log-probability; it determines the conditional relationship
between two species, i.e. the probability that they will be found together,
after controlling for the other species in the network (Figure 2A, Figure 2B).
For example, if $\beta_{ij} = +2$, then each species' odds of occurrence would be
$e^2$ times higher when the other one is present (as compared with otherwise
equivalent sites). The relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur. As a result, the model tends---all else
equal---to produce assemblages where many positively-associated species pairs
co-occur and few negatively-associated pairs do (just as an ecologist might
expect). When all else is *not* equal (e.g. Figure 1, where the presence of one
competitor is associated with release from another competitor), then predicting
species' overall co-occurrence rates would require normalizing the relative 
probabilities of possible $\vec{y}$ vectors so that they sum to one (Figure 2B).

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a set of independent $\vec{y}$ vectors
indicating which species are present at each site on the landscape). When
Equation 1 can be normalized, one can find exact maximum likelihood estimates for
$\alpha$ and $\beta$ using numerical optimization. Fully-observed networks like 
the ones considered here have unimodal likelihood surfaces 
[@murphy_machine_2012], so these estimates will represent the global optimum.

When the number of species is larger than about 30, noralizing Equation 1 
can become intractable, and researchers will either need to approximate it [@lee_learning_2012] or approximate its gradient [@harris_estimating_2016] if
they want to fit a Markov network model. For the analyses presented here, I used 
the rosalia package [@harris_rosalia_2015] for the R programming language
[@r_core_team_r_2015] to calculate $p(\vec{y}; \alpha, \beta)$ and its gradients
exactly [@murphy_machine_2012]; the package passes these functions to the "BFGS" 
method in R's general-purpose optimizer, which finds values for $\alpha$ and 
$\beta$.

\noindent \textbf{Simulated landscapes.}
I simulated several sets of landscapes using known parameters to evaluate
different statistical methods' performance. The first set of landscapes included
the three competing species shown in Figure 1. For each of 1000 replicates, I
generated a landscape with 100 sites by sampling from a probability distribution
defined by the figure's interaction coefficients (Appendix 1). Each of the
methods described below was then evaluated on its ability to correctly infer
that the two shrub species competed with one another, despite their frequent
co-occurrence in unshaded areas.

### 2016-02-12 Harris_2015/manuscript-materials/draft.tex
clarify optimization

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a set of independent \(\vec{y}\) vectors indicating which species are
present at each site on the landscape). When Equation 1 can be
normalized, one can find exact maximum likelihood estimates for
\(\alpha\) and \(\beta\) using numerical optimization. Fully-observed
networks like the ones considered here have unimodal likelihood surfaces
(Murphy 2012), so these estimates will represent the global optimum. I
used the rosalia package (Harris 2015a) for the R programming language
(R Core Team 2015) to calculate \(p(\vec{y}; \alpha, \beta)\) and its
gradients (Murphy 2012); the package passes these functions to the
``BFGS'' method in R's general-purpose optimizer, which then estimates
the Markov network parameters.

\noindent \textbf{Simulated landscapes.} I simulated several sets of
landscapes using known parameters to evaluate different statistical
methods' performance. The first set of landscapes included the three
competing species shown in Figure 1. For each of 1000 replicates, I
generated a landscape with 100 sites by sampling from a probability
distribution defined by the figure's interaction coefficients (Appendix
1). Each of the methods described below was then evaluated on its
ability to correctly infer that the two shrub species competed with one
another, despite their frequent co-occurrence.

@@

Despite their limitations, Markov networks have enormous potential to
improve ecological understanding. In particular, they make many fewer
errors than existing approaches, and can make precise statements about
the conditions where indirect interactions will overwhelm direct ones.
They also provide a simple answer to the question of how competition
should affect a species' overall prevalence, which has important
implications for community-level modeling (Strong et al. 1984).
Specifically, Equation 1 can be used to calculate the expected
prevalence of a species in the absence of biotic influences as
\(e^{\alpha_i}/(e^{0} + e^{\alpha_i})\). Competition's effect on
prevalence can then be estimated by comparing this value with the
observed prevalence (e.g.~comparing Figure 2D with Figure 2C). This
novel quantitative result conflicts with most of our null models, which
unreasonably assume that prevalence would be the exactly same in the
absence of competition as it is in the observed data (Roughgarden 1983).

>>>>

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a set of independent \(\vec{y}\) vectors indicating which species are
present at each site on the landscape). When Equation 1 can be
normalized, one can find exact maximum likelihood estimates for
\(\alpha\) and \(\beta\) using numerical optimization. Fully-observed
networks like the ones considered here have unimodal likelihood surfaces
(Murphy 2012), so these estimates will represent the global optimum.

When the number of species is larger than about 30, noralizing Equation
1 can become intractable, and researchers will either need to
approximate it (Lee and Hastie 2012) or approximate its gradient
({\textbf{???}}). Such approximations are beyond the scope of this
paper, however. For the analyses presented here, I used the rosalia
package (Harris 2015a) for the R programming language (R Core Team 2015)
to calculate \(p(\vec{y}; \alpha, \beta)\) and its gradients exactly
(Murphy 2012); the package passes these functions to the ``BFGS'' method
in R's general-purpose optimizer, which finds values for \(\alpha\) and
\(\beta\).

\noindent \textbf{Simulated landscapes.} I simulated several sets of
landscapes using known parameters to evaluate different statistical
methods' performance. The first set of landscapes included the three
competing species shown in Figure 1. For each of 1000 replicates, I
generated a landscape with 100 sites by sampling from a probability
distribution defined by the figure's interaction coefficients (Appendix
1). Each of the methods described below was then evaluated on its
ability to correctly infer that the two shrub species competed with one
another, despite their frequent co-occurrence in unshaded areas.

@@

Despite their limitations, Markov networks have enormous potential to
improve ecological understanding. In particular, they make many fewer
errors than existing approaches, and can make precise statements about
the conditions where indirect interactions will overwhelm direct ones.
They also provide a simple answer to the question of how competition
should affect a species' overall prevalence, which has important
implications for community-level modeling (Strong et al. 1984).
Specifically, Equation 1 can be used to calculate the expected
prevalence of a species in the absence of biotic influences as
\(e^{\alpha_i}/(e^{0} + e^{\alpha_i})\). Competition's effect on
prevalence can then be estimated by comparing this value with the
observed prevalence (e.g.~comparing Figure 2D with Figure 2C). This
novel quantitative result undermines most of our null models, which
unreasonably assume that prevalence would be the exactly same in the
absence of competition as it is in the observed data (Roughgarden 1983).
